Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-09-18 paper/paperV2.tex
Added V2

\begin{abstract}
The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence. Combined, they allow the prediction of important affective dimensions of the soundscape such as its pleasantness. However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient to predict perceived pleasantness. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes then allows the training and validate the proposed estimation procedure.

@@

\section{Introduction}
\label{sec:intro}
In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the possible positive influence of some of its components.

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In this paper, we will focus on the first affective dimension which is pleasantness. Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. However, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost. Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research. Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using the variations of the measured sound level only. Because these indicators are not representative of the composition of the sound scene, the model underperforms when compared to the predictions computed using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. In order to tackle this issue, the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of data with task-specific annotations are needed for these architectures to learn to extract relevant information from recordings. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with perceptual activity labels. We propose in this paper to study the potential of the use of simulated corpora~\cite{lafay2016, salamon2017} for such purpose.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree of accuracy.
\end{enumerate}

First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings in order to validate the relevance of the proposed protocol for \textit{in situ} sound environments. Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.

\section{Perception of simulated sound scenes}
\label{sec:data}

\subsection{Corpus generation}
\label{sec:data_corp}

A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted. This duration is chosen to be perceptually relevant for the perceptual evaluation of sound environments. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The extracts are then subjectively annotated in terms of the following properties:

@@

The 45~s sound scenes are then replicated using the  \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

@@

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. To do so, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions with parameters identified from annotations of recordings,
\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
\end{itemize}

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Since voice events consist mostly of read English recordings, their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.

In order to gather estimates of the perceptual attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted. Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on the application of pleasantness prediction, including other dimensions can be useful to verify the relevance of simulated data in relation to previous work on real scenarios.

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

@@

\subsection{Corpus validation}
\label{sec:data_val}

@@

\section{Methods}
\label{sec:methods}

\subsection{Acoustical indicators for soundscape description}
\label{sec:methods_inds}

Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters. These indicators can be used to develop baseline physical models of pleasantness in the design of a deep learning architecture for sound source recognition.

@@

\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f1, t)}{df1dt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.

@@

\begin{align}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
\end{align}

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceptual time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.

@@

\subsection{Source presence detection using deep learning}
\label{sec:methods_deep}

Considering machine learning techniques to detect the presence of source requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
\begin{itemize}
\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
\end{itemize}

The generation procedure is the same for both datasets, and is equivalent to that of the 75 scenes in the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}. Only traffic, human voices and birds sources compose the sound scenes. To generate the scenes, it is important to ensure a complete independance of the training and test set. For that, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception. The spectral blocks of dimension 29x8 are then processed independently in this study.

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha_{opt}, \beta_{opt})$.

>>>>

\begin{abstract}

The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure. The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.

@@

\section{Introduction}
\label{sec:intro}

%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/EC European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for noise reduction planning purposes. In this context, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as the "acoustic environment as perceived or experienced and/or understood by people, in context" provides a shift from the noise-only paradigm by considering the possible positive influence of some the sound environment's components.

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}, by identifying its composition in terms of sources of interest from acoustic measurements at a reasonable cost. Classical monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment. Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}. While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline. Indeed, it requires large transmission bandwitdh and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved. Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information. This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}. In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes. Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomiesn it demonstrates that the information content of third ocrave spectra is sufficient to identify sound sources.

The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}. Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}. The perceived time of presence presents the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which is a growing interest in the machine listening community~\cite{mesaros2017}. In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}. Several architectures may be used to such aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}. Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}. Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity. It is possible to manually annotate each recording in the corpus. However, this process can be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources. As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes. By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy.
\end{enumerate}

First, a procedure for simulating sound scenes corpora is designed in Section~\ref{sec:data} and a listening test is conducted to justify its relevance in the context of a study on perceived quality. A large corpus is constructed and used in Section~\ref{sec:pred} to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales. The resulting time of presence are further applied in Section~\ref{sec:app} to the prediction of pleasantness in urban environments, in comparison to existing models from perceptual and physical indicators.

\section{Corpus}
\label{sec:data}

\subsection{Generation procedure}
\label{sec:data_corp}

Considering machine learning techniques to detect the presence of sources requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed. This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality~\cite{lavandier2006}. The corresponding database of isolated is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. \textit{simScene} generates original scenarios using the following information:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
\end{itemize}

Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. To do so, a corpus of 74 recordings obtained in~\cite{aumond2017} is taken as reference. The recordings include 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. The statistics on background and event properties needed for \textit{simScene} generation are extracted conditionally on the ambiance from annotations of the corpus presented in~\cite{gloaguen2017}, which include:

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with random samples from the database of isolated occurences to simulate sound scenes. At the end of the simulation process each sound scene is associated to a random sound level, which is conditioned to the ambiance according to typical sound levels in urban context.

The simulated deep learning corpus is composed of two subsets:
\begin{itemize}
\item The development set, made of 400 scenes of 45~s each (total duration 5 hours), which is used during the training process,
\item The evaluation set, made of 200 scenes of 45~s each (total of 2.5 hours), which is used to compute the performance of the trained model and its generalization capabilities.
\end{itemize}

The generation procedure is the same for both datasets. Though, it is important in deep learning applications to ensure a complete independance of the training and test set. To do so, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.

An experiment in the form a listening test is conducted in order to validate the perceptual correspondance of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.

Given the aim of this test, a separate corpus is constructed. First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}. Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.

@@

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The annotations obtained in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.\\

During the test, participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. The perceived time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on perception is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used by all participants. The calibration of the headphone was carried out in a free field situation (pink noise through a Genelec 1031A loudspeaker in front of the head) and consisted in characterizing the relationship between voltage at the headphone terminal and the sound pressure at the entrance of a blocked ear canal. To do so, small DPA 4060 microphones have been set at the entrance of the ear canals of a human subject~\cite{moller1992}. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listener.

@@

\subsection{Simulated corpus validation}
\label{sec:data_val}

@@

\subsection{Acoustical indicators for soundscape description}
\label{sec:data_inds}

Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters. These indicators may be used to automatically annotate sound scenes in the design of a deep learning architecture for sound source recognition.

@@

\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f_1=31.5Hz}^{f_1=16kHz}\lvert\frac{d^2L(f_1, t)}{df_1dt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. This indicator represents the variations in both the time and frequency dimensions to highlight sources of interest. For example, bird activity is characterized by narrow-band energy with fast paced variations in time, which translates into high $TFSD$ values in the corresponding frequency range. As a baseline indicator for source activity, the $TFSD$ is computed for the 4kHz band and 125ms measurements ($TFSD_{4kHz(1/8s)}$), and for the 500Hz band with 1s measurements ($TFSD_{500Hz, 1s}$) as estimates of the contribution of birds and voices respectively.

@@

\begin{align}\label{eq:ts}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta L_{s}(t, f)\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta L_{s}(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, \hat T_s(\alpha, \beta)\right)
\end{align}

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $N_s$ is the number of sources in the taxonomy and equals 3 in this study, $t$ is the time frame and $T_{s, p}$ corresponds to the perceived time of presence assessments averaged per scene. This indicator evaluates the presence or absence of a given source in a small time frame. First, frequency bands for which the source is emergent by more than the $\alpha$ threshold value are isolated. Then, the source is considered present if the emergence for these bands is on average greater than the $\beta$ threshold. A time of presence estimation is obtained for each source by averaging along the $N_t$ time frames composing the scene. The optimal threshold values are optimized once in this study on the whole perceptual experiment corpus, and found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$.

@@

The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception. The analysis is performed using arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown. Thus, these 6 scenes are excluded of the study. Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.

Table~\ref{tab:physc} shows the Pearson's correlation coefficients between computed physical indicators and perceptual assessments. First, all global sound level indicators correlate well (r>0.9) with the perceived overall loudness. Regarding source-specific perceptual parameters, the $L_{eq, s}$ correlates consistently well with the $T_{s, p}$ of corresponding source (r=0.71). Conversely, the emergence $\Delta L_s$ fails to represent the perceived bird activity, and correlations are weak for other sources. The proposed estimates of the time of presence $\hat T_s(\alpha_{opt}, \beta_{opt})$ show strong correlations to their perceived counterparts (r>0.8), though this is expected as $\alpha_{opt}$ and $\beta_{opt}$ are optimized to this aim. They also display good source discrimination properties, as no significant correlation is found between voices and birds, and perceptual assessments of traffic were correlated with those of other sources in Table~\ref{tab:percc}. They are also better predictors than the $TFSD_{500Hz, 1s}$ and $TFSD_{4kHz, 1/8s}$ indicators for voice and bird activity in this corpus. Thus, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ indicator can be used to annotate simulated sound scenes in terms of estimated perceived time of presence for the three sources.

\section{Source time of presence prediction}
\label{sec:pred}

\subsection{Deep learning model for presence prediction}
\label{sec:pred_deep}

A deep learning model is implemented using the Python Pytorch~\cite{paszke2017} framework and trained on the corpus presented in Section~\ref{sec:data_corp} for source time of presence prediction.

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range as the input signal representation. Instead of a regression task where the output is directly the time of presence, a multiple label classification task on 1~s texture frames is preferred as the resulting training procedure is easier. Individual inputs are thus obtained by splitting the resulting spectrograms into texture windows of 1~s duration. The spectral blocks of dimension 29x8 are then processed independently by the model.

Ground truth target outputs are associated to each input frame to train the model. Considering the results discussed in Section~\ref{sec:data_inds}, the $\hat T_s(\alpha_{opt}, \beta_{opt})$ time of presence estimation seems well suited to automatically label the dataset for this task. Binary presence values on 1~s frames obtained during the $\hat T_s(\alpha_{opt}, \beta_{opt})$ computation before averaging in eq.\ref{eq:ts} are thus used as individual weak labels.
